From physical conditions to Hamiltonian and Lagrangian mechanics

### 2021-02-07 Articles/2020-ClassicalMechanicsPhil/Source/classical-phil-article.tex
Adding notes

In Riemannian geometry, we have a canonical volume form $dV = \epsilon \sqrt{|g_{ab}|} dx^1 \wedge dx^2 \wedge ... \wedge dx^n$ where $|g_{ab}|$ is the determinant of the metric tensor as expressed in local coordinates. In component notation, we can express the volume form as $\sqrt{|g_{ab}|}\epsilon_{a_1 ... a_n}$, where $\epsilon_{a_1 ... a_n}$ is the Levi-Civita symbol. We can then express any differential density $\nu = \hat{\rho} dV$ as the product of the volume density times a scalar function $\hat{\rho} \equiv \rho \ \epsilon \sqrt{|g_{ab}|}$. At first glance, this expression misleadingly suggests that the density $\hat{\rho}$ is coordinate-independent: it is a scalar quantity with the correct dimensions of mass over volume. However, on physical grounds, we require that the numerical value of $\hat{\rho} $ equals the value of $\rho$. This is only possible if the determinant of $g_{ab}$ equals one. Hence, we must define $\hat{\rho} $ in a special coordinate system, namely one where---at least locally---$|g_{ab}| = 1$. For this reason, $\hat{\rho} $ remains coordinate-dependent. 
%Fortunately, we can always find a set of coordinates that, at least locally, makes $|g_{ab}|$ constant. Since the square root of the determinant $|g_{ab}|$ is the Jacobian of the transformation to those coordinates, $|g_{ab}|$ will be constant only when this Jacobian equals one.

@@

In response, we will defend two points. First, we have in fact already in Section~\ref{motion} seen a natural interpretation of both the linear relationship between momentum and velocity and the quadratic Hamiltonian form of equation~\ref{6.2}. Both follow from imposing the motion equivalence condition. In equation 5, we recovered the linear relationship between conjugate momentum and velocity, up to the addition of an arbitrary function (which is connected to the gauge-redundancy of the conjugate momentum). Just as Curiel does, we then used this relationship to derive a quadratic form for the Hamiltonian (Equation 7). Imposing motion equivalence thereby allows us to interpret this quadratic form as the Hamiltonian for massive particles under potential forces. It is not an ad hoc condition imposed on Hamiltonian systems. Instead, it arises in all contexts where motion equivalence holds, i.e. where the trajectory of the system in physical space-time is necessary and sufficient for its evolution in state space.

@@

Secondly, we believe there are some physically-meaningful non-regular Hamiltonian systems. These include models of a classical photon. In general, Hamiltonian systems outside the hyper-regular domain correspond to cases where we cannot express the density of states in terms of position and velocity. These are cases where the weak equivalence condition (equation~\ref{weak_equivalence}) fails. Nevertheless, we can still define a distribution over the phase space points and consider its evolution. We thereby view the broader scope of Hamiltonian mechanics as an advantage, rather than a disadvantage. This interpretation is compatible with the claim that many---perhaps even most---non-regular Hamiltonian systems are physically uninteresting.
%[To do: briefly expand on this and connect to Curiel page 309-310 discussion of a similar system]

@@

More broadly, it seems that Curiel simply privileges a different definition of `classical system' than the one we adopt here. His definition takes satisfying certain kinematical constraints as fundamental. In contrast, we privilege a system having a unique solution, i.e. being deterministic and reversible. Curiel acknowledges that only hyper-regular Lagrangians have unique solutions whereas Hamiltonian systems have unique solutions even outside of the hyper-regular domain \parencites*[291, 308]{Curiel}, but he argues that this is actually an advantage of Lagrangian mechanics, rather than a disadvantage. On his view, the unique solutions of Hamiltonian mechanics outside the hyper-regular domain is precisely why ``one must impose the quadratic form of the Hamiltonian as an \textit{ad hoc} condition" \parencites*[308]{Curiel}. In contrast, he argues that Lagrangian mechanics precisely delimits the proper domain of abstract classical systems, namely those that satisfy the proper kinematical constraints (e.g. $v^i = \dot(x)^i$). Although Lagrangian systems do not admit unique solutions outside the hyper-regular domain, these non-unique solutions will allegedly still satisfy these kinematical constraints \parencites*[311]{Curiel}. Curiel therefore views these non-regular, Lagrangian but non-Hamiltonian systems as physically-meaningful classical systems.

Yet, it is difficult to understand what a Lagrangian system outside the hyper-regular domain could mean. For outside this domain, the Euler--Lagrange equations do not have a single, unique solution. As an example, consider the case where the Lagrangian $L$ equals zero. This Lagrangian corresponds to a Hamiltonian that also equals zero, entailing that $dH = 0 $. By Hamilton's equations, the derivatives of the position and momenta are zero, indicating that this system is static. For this ``system," all paths have the same action, so all paths are possible. This gives an initial probability distribution over the states, but it does not specify how this initial distribution will evolve into a future distribution of states. In other words, the Lagrangian alone does not supply a probability distribution on the possible evolutions. In contrast,  a Hamiltonian defines a unique solution (i.e. a unique evolution in phase space) even when outside the hyper-regular domain (i.e. when the Hessian is zero).\footnote{In a footnote, Barrett conjectures that outside the hyper-regular domain, Hamiltonian and Lagrangian formulations ``will be inequivalent according to any reasonable standard of equivalence. There is no natural way to translate a non-hyper-regular model of the one theory into a model of the other" \parencites*[1179]{Barrett2}.}

@@

>>>>

In Riemannian geometry, we have a canonical volume form $dV = \epsilon \sqrt{|g_{ab}|} dx^1 \wedge dx^2 \wedge ... \wedge dx^n$ where $|g_{ab}|$ is the determinant of the metric tensor as expressed in local coordinates. In component notation, we can express the volume form as $\sqrt{|g_{ab}|}\epsilon_{a_1 ... a_n}$, where $\epsilon_{a_1 ... a_n}$ is the Levi-Civita symbol. We can then express any differential density $\nu = \hat{\rho} dV$ as the product of the volume density times a scalar function $\hat{\rho} \equiv \rho \ \epsilon \sqrt{|g_{ab}|}$. At first glance, this expression misleadingly suggests that the density $\hat{\rho}$ is coordinate-independent: it is a scalar quantity with the correct dimensions of mass over volume. However, on physical grounds, we require that the numerical value of $\hat{\rho} $ equals the value of $\rho$. This is only possible if the determinant of $g_{ab}$ equals one. Hence, we must define $\hat{\rho} $ in a special coordinate system, namely one where---at least locally---$|g_{ab}| = 1$. For this reason, $\hat{\rho} $ remains coordinate-dependent. [XXX: more than locally.... remove locally]
%Fortunately, we can always find a set of coordinates that, at least locally, makes $|g_{ab}|$ constant. Since the square root of the determinant $|g_{ab}|$ is the Jacobian of the transformation to those coordinates, $|g_{ab}|$ will be constant only when this Jacobian equals one.

@@

In response, we will defend two points. First, we have in fact already in Section~\ref{motion} seen a natural interpretation of both the linear relationship between momentum and velocity and the quadratic Hamiltonian form of equation~\ref{6.2}. Both follow from imposing the motion equivalence condition. In equation 5, we recovered the linear relationship between conjugate momentum and velocity, up to the addition of an arbitrary function (which is connected to the gauge-redundancy of the conjugate momentum) [XXX: and the magnetic force -- ]. Just as Curiel does, we then used this relationship to derive a quadratic form for the Hamiltonian (Equation 7). Imposing motion equivalence thereby allows us to interpret this quadratic form as the Hamiltonian for massive particles under potential forces. It is not an ad hoc condition imposed on Hamiltonian systems. Instead, it arises in all contexts where motion equivalence holds, i.e. where the trajectory of the system in physical space-time is necessary and sufficient for its evolution in state space.

@@

Secondly, we believe there are some physically-meaningful non-regular Hamiltonian systems. These include models of a classical photon. [XXX: a massless particle that moves at constant speed] In general, Hamiltonian systems outside the hyper-regular domain correspond to cases where we cannot express the density of states in terms of position and velocity. These are cases where the weak equivalence condition (equation~\ref{weak_equivalence}) fails. Nevertheless, we can still define a distribution over the phase space points and consider its evolution. We thereby view the broader scope of Hamiltonian mechanics as an advantage, rather than a disadvantage. This interpretation is compatible with the claim that many---perhaps even most---non-regular Hamiltonian systems are physically uninteresting.
%[To do: briefly expand on this and connect to Curiel page 309-310 discussion of a similar system]

@@

More broadly, it seems that Curiel simply privileges a different definition of `classical system' than the one we adopt here. His definition takes satisfying certain kinematical constraints as fundamental. [XXX: state defined at equal time, velocity is not] In contrast, we privilege a system having a unique solution, i.e. being deterministic and reversible. Curiel acknowledges that only hyper-regular Lagrangians have unique solutions whereas Hamiltonian systems have unique solutions even outside of the hyper-regular domain \parencites*[291, 308]{Curiel}, but he argues that this is actually an advantage of Lagrangian mechanics, rather than a disadvantage. On his view, the unique solutions of Hamiltonian mechanics outside the hyper-regular domain is precisely why ``one must impose the quadratic form of the Hamiltonian as an \textit{ad hoc} condition" \parencites*[308]{Curiel}. In contrast, he argues that Lagrangian mechanics precisely delimits the proper domain of abstract classical systems, namely those that satisfy the proper kinematical constraints (e.g. $v^i = \dot{x}^i$). Although Lagrangian systems do not admit unique solutions outside the hyper-regular domain, these non-unique solutions will allegedly still satisfy these kinematical constraints \parencites*[311]{Curiel}. Curiel therefore views these non-regular, Lagrangian but non-Hamiltonian systems as physically-meaningful classical systems.

Yet, it is difficult to understand what a Lagrangian system outside the hyper-regular domain could mean. [XXX: typically, non-regular lagrangian are broken into regions where they give one solution] For outside this domain, the Euler--Lagrange equations do not have a single, unique solution. As an example, consider the case where the Lagrangian $L$ equals zero. This Lagrangian gives rise to a Hamiltonian that also equals zero, entailing that $H = 0 $. By Hamilton's equations, the derivatives of the position and momenta are zero, indicating that this system is static. For this ``system," all paths have the same action, so all paths are possible. This gives an initial probability distribution over the states, but it does not specify how this initial distribution will evolve into a future distribution of states. In other words, the Lagrangian alone does not supply a probability distribution on the possible evolutions. In contrast,  a Hamiltonian defines a unique solution (i.e. a unique evolution in phase space) even when outside the hyper-regular domain (i.e. when the Hessian is zero).\footnote{In a footnote, Barrett conjectures that outside the hyper-regular domain, Hamiltonian and Lagrangian formulations ``will be inequivalent according to any reasonable standard of equivalence. There is no natural way to translate a non-hyper-regular model of the one theory into a model of the other" \parencites*[1179]{Barrett2}.}

@@

[XXX: missing bit about changing assumption to quantum]

Why we prefer this view:
\begin{itemize}
	\item Explains why statistical mechanics is best done on phase space and the role of the symplectic form in counting states for each degree of freedom (two for the price of one)
	\item Clarifies the physical role of each mathematical element (no hiding in mathematical jargon)
	\item Allows for a more geometrical explanation of the action principle
	\item Gives a simple understanding of gauge transformations
	\item Change of one assumption recovers quantum mechanics (clarifies the boundary between the two) refer to long paper
\end{itemize}
